---
title: "Biodiversity in Tennessee, 2017"
author: "Alex Austin"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 10
    fig_height: 7
    toc: yes
    number_sections : yes
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning=FALSE,out.width='1000px', dpi=200)
```

```{r}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(zoo)
library(RColorBrewer)
library(wesanderson)
library(choroplethr)
library(choroplethrMaps)
library(neuralnet)
library(h2o)
```

#Introduction
This dataset is composed of 2 csv files :

* one describing the national US parks locations
* one describing the presence and status of species : species ID, park name, taxonomic information, scientific name, one or more common names, record status, occurrence, nativeness, abundance, seasonality (season and nature of presence in park), and conservation status

It may be interesting to look a the distribution per park, and see if there are species in danger in a given partk/location


#Data
```{r}
species<-read.csv('input/species.csv',sep=',')
parks<-read.csv('input/parks.csv',sep=',')
colnames(species)
colnames(parks)
```

We can see that there is a common column `Park.Name` with the 2 files so we can merge the 2 files for later use.
We can also see that the `Conservation.Status` has an empty string, meaning the Status is safe (assumption made given the toal number of these cases). So we can rename the empty string for better ocmprehension.

```{r}
status<-function(x){
  if(x==''){return('Safe')}
  else {return(as.character(x))}
}
species$Status<-sapply(species$Conservation.Status,status)
```

```{r}
merged<-merge(species, parks,by="Park.Name")
#head(merged)
```
<hr>

#Distribution of species
As observed when looking at the dataframes, we might disentangle the 2 types of `Statuses` : `Safe` and `others`
```{r}
Cat<-as.data.frame(merged %>% dplyr::group_by(Category,Status) %>% dplyr::summarise(number = n()) %>% arrange(-number))
#define the colorPalette

colourCount = length(unique(Cat$Status))
getPalette = colorRampPalette(wes_palette("Zissou"))
p1<-ggplot(filter(Cat,Status=='Safe')) + 
  geom_bar(aes(x=reorder(Category,-number),y=number,fill=factor(Status)),stat="identity") + 
  scale_fill_manual(values = getPalette(colourCount)) + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  xlab('') + ylab('Number of species') + 
  theme(legend.title=element_blank(),legend.position='top',legend.text=element_text(size=8),legend.key.size = unit(.4, "cm"))

colourCount2 = length(unique(Cat$Status))
getPalette2 = colorRampPalette(wes_palette("FantasticFox"))
p2<-ggplot(filter(Cat,Status!='Safe')) + 
  geom_bar(aes(x=reorder(Category,-number),y=number,fill=factor(Status)),stat="identity") + 
  scale_fill_manual(values = getPalette2(colourCount2)) + theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  xlab('') + ylab('Number of species') + 
  theme(legend.title=element_blank(),legend.position='top',legend.text=element_text(size=8),legend.key.size = unit(.4, "cm"))

grid.arrange(p1,p2,ncol=2)
```

#Distribution of species per park and Status
As seen we the plot above, we might disentangle the 2 types of `Statuses` : `Safe` and `others`
```{r}
Cat2<-as.data.frame(merged %>% 
                      dplyr::group_by(Category,Park.Name,Status) %>% 
                      dplyr::summarise(number = n()) %>% 
                      arrange(-number))

p3<-ggplot(data=filter(Cat2,Status=='Safe'), aes(Category,Park.Name)) + 
  geom_tile(aes(fill = number),colour = "white") + 
  scale_fill_gradient(low = "pink" ,high = "black" ) + 
  theme(legend.position='top',axis.text.y = element_text(size=8),axis.text.x = element_text(size=8,angle=45, hjust=1)) + 
  xlab('') + ylab('')

print(p3)

p4<-ggplot(data=filter(Cat2,Status!='Safe'), aes(Category,Park.Name)) + 
  geom_tile(aes(fill = number),colour = "white") + 
  scale_fill_gradient(low = "pink" ,high = "black" ) + 
  theme(legend.position='top',axis.text.y = element_text(size=8),axis.text.x = element_text(size=8,angle=45, hjust=1)) + 
  xlab('') + ylab('')

print(p4)
```

##Comments

* as we saw before, most of the species are of `Plant` category
* There is a park (_Great Smoky Mountains Nat. Park_) where Insect are a dominant population (`Safe` status)
* The _Mammal_ is somehow a category in danger in several parks

#State Map
The `parks.csv` returns the name and State of the corresponding park. I wanted to use a `choropleth` map.
However the input of this type of map (if I understand well) if : `value,region` where `region` is the state in lower case.
`R` has functions `state.abb`, `state.name` that returns either a State acronym or name. The issue with `parks.csv` is that some park are shared between 3 States. So Idefined a function to return only one State (corresponding with the one having the largest area of the park).

```{r}
convertName<-function(x){
for(i in 1:length(state.abb)){
	if(as.character(x) == state.abb[i]){
		return(state.name[i])
		break
	   }
	   else if(as.character(x) == 'TN, NC') {
	   	return(state.name[42]) #Tennessee
	   	break
	   	}
	   	else if(as.character(x) == 'WY, MT, ID') {
	   	return(state.name[50]) #Wyoming
	   	break
	   	}
	   	else if(as.character(x) == 'CA, NV') {
	   	return(state.name[5]) # cal.
	   	break
	   	}   	
	}	
}
```

For this example, I group-by the data by state only :
```{r}
speciesByState<-as.data.frame(
  merged %>% 
    dplyr::group_by(State) %>% 
    na.omit() %>%
    dplyr::summarise(number = n()) 
  %>% arrange(-number))
speciesByState$State<-sapply(speciesByState$State,convertName)

#aggregate the remaining rows having the same `state`
speciesByState<-aggregate(. ~ State, data=speciesByState, FUN=sum)

#change to lower
speciesByState$State<-sapply(speciesByState$State, tolower)

#change columns name for choropleth
colnames(speciesByState)<-c('region','value')
head(speciesByState)

#map
getPalette = colorRampPalette(wes_palette("Zissou"))
c = StateChoropleth$new(speciesByState)
c$ggplot_scale = scale_fill_manual(name='# of Species',values = getPalette(7))
c$render() + ggtitle("2017 Species by State")
#state_choropleth(speciesByState, title="2017 Species by State", legend="Species")
```

```{r}
#find top species
bySpecies<-as.data.frame(merged %>% dplyr::group_by(Category) %>% dplyr::summarise(number = n()) %>% arrange(-number))
par(mfrow = c(2, 2))
#find top species
bySpecies<-as.data.frame(merged %>% dplyr::group_by(Category) %>% dplyr::summarise(number = n()) %>% arrange(-number))
mylist<-list()
for(i in 1:4){
	ByState<-as.data.frame(merged %>% filter(Category==bySpecies$Category[i]) %>% dplyr::group_by(State) %>% dplyr::summarise(number = n()) %>% arrange(-number))
	#print(head(ByState))
	ByState$State<-sapply(ByState$State,convertName)
	ByState<-aggregate(. ~ State, data=ByState, FUN=sum)
	ByState$State<-sapply(ByState$State, tolower)
	ByState<- ByState %>% dplyr::rename(region=`State`,value=`number`)
	mylist[[i]]<-ByState
	#state_choropleth(ByState, title=paste("Distribution of",as.character(bySpecies$Category[i])), num_colors=9)
}

makeMap<-function(cnt){
  temp<-StateChoropleth$new(mylist[[cnt]])
  temp$ggplot_scale = scale_fill_manual(name='# of Species',values = getPalette(7))
  d<-temp$render() + ggtitle (paste("Distribution of",as.character(bySpecies$Category[cnt])))
  return(d)
}

grid.arrange(
  makeMap(1), makeMap(2), makeMap(3), makeMap(4),
  ncol=2
)
```
<hr>
<strong>History :</strong>

* _version 1 : initial commit_ 
* _version 2 : added choropleth maps, using a not-very elegant coding style_
* _version 3 : updated for new csv files_
* _version 4 : added grid of choropleth maps for the top 4 species_
* _version 5 : reshape sections_
<hr>